# Forest {.unnumbered}

```{r setup}
#| include: false
rm(list = ls()) ; invisible(gc()) ; set.seed(42)
library(tidyverse)
library(terra)
library(sf)
library(leaflet)
theme_set(bayesplot::theme_default())
knitr::opts_chunk$set(
  echo = F, message = F, warning = F, fig.height = 6, fig.width = 8,
  cache = T, cache.lazy = F)
```

We used the Tropical Moist Forest (TMF) product from the European Joint Research Centre (JRC) to asses forest surface evolution between 2001 and 2020 at 30-m and yearly resolutions. We focused on pixels categorized as undisturbed forest in 2001 and assessed if they were (1) undisturbed, (2) degraded or (3) deforested in 2020. We aggregated the resulting map to 0.01-degree by computing the percentage of degraded and deforested pixels.

```{r tmfdata}
#| include: false
#| eval: false
temp_r <- rast("results/data/modis.nc")
area <- st_read("results/limits/limits.shp", quiet=T) %>% 
  st_set_crs(crs(temp_r))
prep_tmf <- function(tile){
  tmf_2001 <- rast(
    paste0("data/JRC_TMF_AnnualChange_v1_", 
           tile,
           "/forobs/products/tmf_v1/AnnualChange/JRC_TMF_AnnualChange_v1_2001_",
           tile,
           ".tif")) %>% crop(area)
    msk <- ifel(tmf_2001 != 1, NA, 1)
    # writeRaster(msk, paste0("results/tmp/tmf_", tile, "_forest_2001.tif"))
    tmf_2020 <- rast(
    paste0("data/JRC_TMF_AnnualChange_v1_", 
           tile,
           "/forobs/products/tmf_v1/AnnualChange/JRC_TMF_AnnualChange_v1_2020_",
           tile,
           ".tif")) %>% crop(area) %>% mask(msk)
    tmf_2020 <- as.factor(ifel(tmf_2020 %in% 1:3, tmf_2020, NA))
    degrad_2020 <- as.numeric(tmf_2020 == 2) %>% 
      aggregate(37, method = "mean")
    # %>% resample(temp_r)
    deforest_2020 <- as.numeric(tmf_2020 == 3) %>% 
      aggregate(37, method = "mean")
    # %>% resample(temp_r)
    writeRaster(degrad_2020, paste0("results/tmp/tmf_", tile, "_degraded_2020.tif"))
    writeRaster(deforest_2020, paste0("results/tmp/tmf_", tile, "_deforested_2020.tif"))
}
lapply(c("SAM_ID28_N0_W80",
         "SAM_ID29_N0_W70",
         "SAM_ID47_N10_W80",
         "SAM_ID48_N10_W70"),
       prep_tmf)
```

## All

```{r deforest2021g}
temp_r <- rast("results/data/modis_anomalies.nc")
area <- st_read("results/area/Guaviare/Guaviare.shp", quiet=T) %>% 
  st_set_crs(crs(temp_r))
deforest <- rast("results/data/tmf_deforested_2020.tif") %>% 
  resample(temp_r) %>% 
  crop(area) %>% 
  mask(area)
ggplot() +
  tidyterra::geom_spatraster(data = deforest) +
  theme_bw() +
  geom_sf(data = area, fill = NA, linewidth = 2) +
  ggtitle("Guiavare: percentage of deforestation in 2020") +
  scale_fill_viridis_c(na.value = NA)
```

```{r degrad2020g}
temp_r <- rast("results/data/modis_anomalies.nc")
area <- st_read("results/area/Guaviare/Guaviare.shp", quiet=T) %>% 
  st_set_crs(crs(temp_r))
degrad <- rast("results/data/tmf_degraded_2020.tif") %>% 
  resample(temp_r) %>% 
  crop(area) %>% 
  mask(area)
ggplot() +
  tidyterra::geom_spatraster(data = degrad) +
  theme_bw() +
  geom_sf(data = area, fill = NA, linewidth = 2) +
  ggtitle("Guiavare: percentage of degradation in 2020") +
  scale_fill_viridis_c(na.value = NA)
```

```{r histg}
data <- data.frame(deforest = values(deforest),
                   degrad = values(degrad))
names(data) <- c("deforested", "degraded")
data %>% 
  gather(variable, value) %>% 
  na.omit() %>% 
  ggplot(aes(value)) +
  geom_histogram() +
  theme_bw() +
  facet_wrap(~ variable ~ (value > 0.05), 
             scales = "free_y", labeller = label_both)
```

```{r tmfcor}
data <- as.data.frame(deforest, xy = TRUE) %>% 
  rename(deforested = Dec2020) %>% 
  left_join(as.data.frame(degrad, xy = TRUE)) %>% 
  rename(degraded = Dec2020) %>% 
  na.omit() %>% 
  sample_n(1000)
cor1 <- pgirmess::correlog(data.frame(data$x, data$y), data$deforested,
                   method = "Moran", nbclass = 30) %>% 
  as.data.frame() %>% 
  mutate(variable = "deforested")
cor2 <- pgirmess::correlog(data.frame(data$x, data$y), data$degraded,
                   method = "Moran", nbclass = 30) %>% 
  as.data.frame() %>% 
  mutate(variable = "degraded")
bind_rows(cor1, cor2) %>% 
  ggplot(aes(x = dist.class*100, y = coef, col = variable)) + 
  geom_point(aes(alpha = p.value < 0.01)) + geom_line() +
  geom_hline(yintercept = 0) +
  theme_bw() +
  xlab("Distance (km)") + ylab("Moran\'s I") +
  scale_x_log10()
```

## Non null

```{r deforest2021}
temp_r <- rast("results/data/modis_anomalies.nc")
area <- st_read("results/area/Guaviare/Guaviare.shp", quiet=T) %>% 
  st_set_crs(crs(temp_r))
deforest <- rast("results/data/tmf_deforested_2020.tif") %>% 
  resample(temp_r) %>% 
  crop(area) %>% 
  mask(area)
deforest <- ifel(deforest > 0, deforest, NA)
ggplot() +
  tidyterra::geom_spatraster(data = deforest) +
  theme_bw() +
  geom_sf(data = area, fill = NA, linewidth = 2) +
  ggtitle("Guiavare: percentage > 0 of deforestation in 2020") +
  scale_fill_viridis_c(na.value = NA)
```

```{r degrad2020}
temp_r <- rast("results/data/modis_anomalies.nc")
area <- st_read("results/area/Guaviare/Guaviare.shp", quiet=T) %>% 
  st_set_crs(crs(temp_r))
degrad <- rast("results/data/tmf_degraded_2020.tif") %>% 
  resample(temp_r) %>% 
  crop(area) %>% 
  mask(area)
degrad <- ifel(degrad > 0.05, degrad, NA)
ggplot() +
  tidyterra::geom_spatraster(data = degrad) +
  theme_bw() +
  geom_sf(data = area, fill = NA, linewidth = 2) +
  ggtitle("Guiavare: percentage > 5% of degradation in 2020") +
  scale_fill_viridis_c(na.value = NA)
```

```{r hist}
data <- data.frame(deforest = values(deforest),
                   degrad = values(degrad))
names(data) <- c("deforested", "degraded")
data %>% 
  gather(variable, value) %>% 
  na.omit() %>% 
  ggplot(aes(value)) +
  geom_histogram() +
  theme_bw() +
  facet_wrap(~ variable, 
             scales = "free_y", labeller = label_both)
```

```{r tmfcorz}
data <- as.data.frame(deforest, xy = TRUE) %>% 
  rename(deforested = Dec2020) %>% 
  left_join(as.data.frame(degrad, xy = TRUE)) %>% 
  rename(degraded = Dec2020) %>% 
  na.omit()
cor1 <- pgirmess::correlog(data.frame(data$x, data$y), data$deforested,
                   method = "Moran", nbclass = 30) %>% 
  as.data.frame() %>% 
  mutate(variable = "deforested")
cor2 <- pgirmess::correlog(data.frame(data$x, data$y), data$degraded,
                   method = "Moran", nbclass = 30) %>% 
  as.data.frame() %>% 
  mutate(variable = "degraded")
bind_rows(cor1, cor2) %>% 
  ggplot(aes(x = dist.class*100, y = coef, col = variable)) + 
  geom_point(aes(alpha = p.value < 0.01)) + geom_line() +
  geom_hline(yintercept = 0) +
  theme_bw() +
  xlab("Distance (km)") + ylab("Moran\'s I") +
  scale_x_log10()
```
